---
title: "Part2"
output: html_document
---

Setting up R packages
```{r}
# install required packages only if missing
packages <- c("DBI", "RSQLite")
for (pkg in packages){
  if(!requireNamespace(pkg, quietly = TRUE)){
    install.packages(pkg)
  }
}
library(DBI)
library(RSQLite)
```

Init DB connection
```{r}
# connect to SQLite
con <- dbConnect(SQLite(), "dataverse_db.sqlite")

```

Loading all relevant data into DB
```{r}
# load zipped csv & bz2 files into SQLite
folder <- "../../Data"
# create list containing the files
csv_files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
bz2_files <- list.files(folder, pattern = "\\.bz2$", full.names = TRUE)

# import each file into the db of type .csv
for (file in csv_files) {
  table_name <- tools::file_path_sans_ext(basename(file))
message("Importing CSV: ", file)
df <- read.csv(file)
dbWriteTable(con, table_name, df, overwrite = TRUE)
}

# import each file into the db of type .bz2
for (file in bz2_files) {
  table_name <- tools::file_path_sans_ext(basename(file))
message("Importing BZ2: ", file)
df <- read.csv(file)
dbWriteTable(con, table_name, df, overwrite = TRUE)
}
```



Question a: What are the best times and days of the week to minimise delays each year?

A solution to solve this would be to take the average delays, positive or negative (early), per time and day and return a value to return the given results. The data consists out of the following types of delays: arrival, departure, carrier (consists out of crew, maintenance, cleaning, baggage, ...), weather, national airspace system (eg ATC issues, ...), security and late aircraft delays. [@cite https://www.aspm.faa.gov/aspmhelp/index/Types_of_Delay.html#NAS_Delay] We will exclude any cancelled and diverted flights, as there cant be attributed any delays. We will query the database over the provided years 2004-8.The resulting findings, will reflect the typical evaluation used by the FAA and airlines to evaluate the average delays.
```{r}
ex <- dbGetQuery(con, "SELECT 
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	f.DayOfWeek,
	AVG(f.DepDelay) AS avg_depDelay,
	AVG(f.ArrDelay) AS avg_arrDelay,
	AVG(f.CarrierDelay) AS avg_carrierDelay,
	AVG(f.WeatherDelay) AS avg_weatherDelay,
	AVG(f.NASDelay) AS avg_nasDelay,
	AVG(f.SecurityDelay) AS avg_securityDelay,
	AVG(f.LateAircraftDelay) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY f.DayOfWeek
) AS agg
Order BY agg.DayOfWeek")
print(ex)
```

If however we truly want to see the positive delays [@cite https://catsr.vse.gmu.edu/pubs/XuMultiFactorModelAirportDelaysTRBv6.pdf, p 3 Ch "Definition and previous research"], thus excluding any so called negative events such as early arrivals, departures, ..., thus eliminating any factors which may reduce the average times. As a result, we will get the real average delay times which occurs. This will truly reflect the times experienced by passengers on delays when they happen.

```{r}
positveDelays <- dbGetQuery(con, "SELECT 
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	f.DayOfWeek,
	AVG(
		CASE 
			WHEN f.DepDelay > 0
			THEN f.DepDelay
			ELSE NULL
		END
		) AS avg_depDelay,
	AVG(
		CASE
			WHEN f.ArrDelay > 0
			THEN f.ArrDelay
			ELSE NULL
		END
		) AS avg_arrDelay,
	AVG(
		CASE
			WHEN f.CarrierDelay > 0
			THEN f.CarrierDelay
			ELSE NULL
		END
		) AS avg_carrierDelay,
	AVG(
		CASE
			WHEN f.WeatherDelay > 0
			THEN f.WeatherDelay
			ELSE NULL
		END
		) AS avg_weatherDelay,
	AVG(
		CASE
			WHEN f.NASDelay > 0
			THEN f.NASDelay
			ELSE NULL
		END
		) AS avg_nasDelay,
	AVG(
		CASE
			WHEN f.SecurityDelay > 0
			THEN f.SecurityDelay
			ELSE NULL
		END
		) AS avg_securityDelay,
	AVG(
		CASE
			WHEN f.LateAircraftDelay > 0
			THEN f.LateAircraftDelay
			ELSE NULL
		END
		) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY f.DayOfWeek
) AS agg
Order BY agg.DayOfWeek")
print(positveDelays)

```

Looking at the results from both queries, we can see that Saturday would be the statistically most optimal day to travel to avoid delays during the period of 2004-2008. Evaluating the second most optimal day however proves to be harder, as the average delay calculation differs from the results provided by looking at only positive delays, showing the existing realities present.




Disconnect DB

```{r}
dbDisconnect(con)
```


References: https://www.aspm.faa.gov/aspmhelp/index/Types_of_Delay.html#NAS_Delay