---
title: "Part2"
output: html_document
---

Setting up R packages
```{r}
# install required packages only if missing
packages <- c("DBI", "RSQLite")
for (pkg in packages){
  if(!requireNamespace(pkg, quietly = TRUE)){
    install.packages(pkg)
  }
}
library(DBI)
library(RSQLite)
```

Init DB connection
```{r}
# connect to SQLite
con <- dbConnect(SQLite(), "dataverse_db.sqlite")

```

Loading all relevant data into DB
```{r}
# load zipped csv & bz2 files into SQLite
folder <- "../../Data"
# create list containing the files
csv_files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
bz2_files <- list.files(folder, pattern = "\\.bz2$", full.names = TRUE)

# import each file into the db of type .csv
for (file in csv_files) {
  table_name <- tools::file_path_sans_ext(basename(file))
message("Importing CSV: ", file)
df <- read.csv(file)
dbWriteTable(con, table_name, df, overwrite = TRUE)
}

# import each file into the db of type .bz2
for (file in bz2_files) {
  table_name <- tools::file_path_sans_ext(basename(file))
message("Importing BZ2: ", file)
df <- read.csv(file)
dbWriteTable(con, table_name, df, overwrite = TRUE)
}
```



Question a: What are the best times and days of the week to minimise delays each year?

A solution to solve this would be to take the average delays, positive or negative (early), per time and day and return a value to return the given results. The data consists out of the following types of delays: arrival, departure, carrier (consists out of crew, maintenance, cleaning, baggage, ...), weather, national airspace system (eg ATC issues, ...), security and late aircraft delays. [@cite https://www.aspm.faa.gov/aspmhelp/index/Types_of_Delay.html#NAS_Delay] We will exclude any cancelled and diverted flights, as there cant be attributed any delays. We will query the database over the provided years 2004-8.The resulting findings, will reflect the typical evaluation used by the FAA and airlines to evaluate the average delays.
```{r}
ex <- dbGetQuery(con, "SELECT 
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	f.DayOfWeek,
	AVG(f.DepDelay) AS avg_depDelay,
	AVG(f.ArrDelay) AS avg_arrDelay,
	AVG(f.CarrierDelay) AS avg_carrierDelay,
	AVG(f.WeatherDelay) AS avg_weatherDelay,
	AVG(f.NASDelay) AS avg_nasDelay,
	AVG(f.SecurityDelay) AS avg_securityDelay,
	AVG(f.LateAircraftDelay) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY f.DayOfWeek
) AS agg
Order BY agg.DayOfWeek")
print(ex)
```

If however we truly want to see the positive delays [@cite https://catsr.vse.gmu.edu/pubs/XuMultiFactorModelAirportDelaysTRBv6.pdf, p 3 Ch "Definition and previous research"], thus excluding any so called negative events such as early arrivals, departures, ..., thus eliminating any factors which may reduce the average times. As a result, we will get the real average delay times which occurs. This will truly reflect the times experienced by passengers on delays when they happen.

```{r}
positveDelays <- dbGetQuery(con, "SELECT 
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	f.DayOfWeek,
	AVG(
		CASE 
			WHEN f.DepDelay > 0
			THEN f.DepDelay
			ELSE NULL
		END
		) AS avg_depDelay,
	AVG(
		CASE
			WHEN f.ArrDelay > 0
			THEN f.ArrDelay
			ELSE NULL
		END
		) AS avg_arrDelay,
	AVG(
		CASE
			WHEN f.CarrierDelay > 0
			THEN f.CarrierDelay
			ELSE NULL
		END
		) AS avg_carrierDelay,
	AVG(
		CASE
			WHEN f.WeatherDelay > 0
			THEN f.WeatherDelay
			ELSE NULL
		END
		) AS avg_weatherDelay,
	AVG(
		CASE
			WHEN f.NASDelay > 0
			THEN f.NASDelay
			ELSE NULL
		END
		) AS avg_nasDelay,
	AVG(
		CASE
			WHEN f.SecurityDelay > 0
			THEN f.SecurityDelay
			ELSE NULL
		END
		) AS avg_securityDelay,
	AVG(
		CASE
			WHEN f.LateAircraftDelay > 0
			THEN f.LateAircraftDelay
			ELSE NULL
		END
		) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		DayOfWeek,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY f.DayOfWeek
) AS agg
Order BY agg.DayOfWeek")
print(positveDelays)

```

Looking at the results, which I have decided to expose each average sum of delay type per day and the total from both queries, we can see that Saturday would be the statistically most optimal day to travel to avoid delays during the period of 2004-2008. Evaluating the second most optimal day however proves to be harder, as the average delay calculation differs from the results provided by looking at only positive delays, showing the existing realities present.


Question B: 
In the airline business the evaluation of age of an aircraft is not defined by years in service but rather flight hours and pressurization cycles. Unfortunately this data is logged in a planes logbook, which isn't available in the resource provided or publicly available. One could evaluate the flights hours flown within the given data time frame available, however lacking the previous years of data and crucial maintenance cycles administered to the aircraft, would make it very hard to judge if a given aircraft is reaching the end of its life cycle or not and if life extending measures or total overhaul measures have been taken place. 

To answer this question, I will use an arbitrary definition of an old plane as being 20 plus years in service. Using the year date available in the plane-data.csv file and again differentiate between the average total and only positive delays that occurred with the aircraft between the period from 2004-2008, again excluding cancelled and diverted planes.
```{r}
avgDelaysPlaneAge <- dbGetQuery(con, "SELECT 
	agg.plane_age_group,
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	CASE 
		WHEN (f.Year - p.Year) < 20 
		THEN 'newPlane'
		ELSE 'oldPlane'
	END AS plane_age_group,
	AVG(f.DepDelay) AS avg_depDelay,
	AVG(f.ArrDelay) AS avg_arrDelay,
	AVG(f.CarrierDelay) AS avg_carrierDelay,
	AVG(f.WeatherDelay) AS avg_weatherDelay,
	AVG(f.NASDelay) AS avg_nasDelay,
	AVG(f.SecurityDelay) AS avg_securityDelay,
	AVG(f.LateAircraftDelay) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		Year,
		TailNum,
		ArrDelay,
		DepDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f
	LEFT JOIN (
		SELECT tailnum, Year
		FROM 'plane-data'
		) AS p 
	ON f.TailNum = p.tailnum 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY plane_age_group
) AS agg
Order BY agg.plane_age_group")
print(avgDelaysPlaneAge)

```



```{r}
postiveDelaysPlaneAge <- dbGetQuery(con, "SELECT 
	agg.plane_age_group,
	agg.avg_depDelay,
	agg.avg_arrDelay,
	agg.avg_carrierDelay,
	agg.avg_weatherDelay,
	agg.avg_nasDelay,
	agg.avg_securityDelay,
	agg.avg_lateAircraftDelay,
	(
		agg.avg_depDelay +
		agg.avg_arrDelay +
		agg.avg_carrierDelay +
		agg.avg_weatherDelay +
		agg.avg_nasDelay +
		agg.avg_securityDelay +
		agg.avg_lateAircraftDelay 
	) / 7 AS total_avg_delay
FROM (
SELECT 
	CASE 
		WHEN (f.Year - p.Year) < 20 
		THEN 'newPlane'
		ELSE 'oldPlane'
	END AS plane_age_group,
	AVG(
		CASE 
			WHEN f.DepDelay > 0
			THEN f.DepDelay
			ELSE NULL
		END
		) AS avg_depDelay,
	AVG(
		CASE
			WHEN f.ArrDelay > 0
			THEN f.ArrDelay
			ELSE NULL
		END
		) AS avg_arrDelay,
	AVG(
		CASE
			WHEN f.CarrierDelay > 0
			THEN f.CarrierDelay
			ELSE NULL
		END
		) AS avg_carrierDelay,
	AVG(
		CASE
			WHEN f.WeatherDelay > 0
			THEN f.WeatherDelay
			ELSE NULL
		END
		) AS avg_weatherDelay,
	AVG(
		CASE
			WHEN f.NASDelay > 0
			THEN f.NASDelay
			ELSE NULL
		END
		) AS avg_nasDelay,
	AVG(
		CASE
			WHEN f.SecurityDelay > 0
			THEN f.SecurityDelay
			ELSE NULL
		END
		) AS avg_securityDelay,
	AVG(
		CASE
			WHEN f.LateAircraftDelay > 0
			THEN f.LateAircraftDelay
			ELSE NULL
		END
		) AS avg_lateAircraftDelay
FROM ( 
	SELECT 
		Year,
		TailNum,
		ArrDelay,
		DepDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2004.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2005.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2006.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2007.csv'
	UNION ALL 
	SELECT
		Year,
		TailNum,
		DepDelay,
		ArrDelay,
		CarrierDelay,
		WeatherDelay,
		NASDelay,
		SecurityDelay,
		LateAircraftDelay,
		Diverted, 
		Cancelled FROM '2008.csv'
	) AS f
	LEFT JOIN (
		SELECT tailnum, Year
		FROM 'plane-data'
		) AS p 
	ON f.TailNum = p.tailnum 
	WHERE f.Cancelled = 0 
	AND f.Diverted = 0 
Group BY plane_age_group
) AS agg
Order BY agg.plane_age_group")
print(postiveDelaysPlaneAge)

```




Disconnect DB

```{r}
dbDisconnect(con)
```


References: https://www.aspm.faa.gov/aspmhelp/index/Types_of_Delay.html#NAS_Delay